Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-21 paper.tex
Final^2 changes

\subsection{Decoupled Fallback Statistics}
\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@

To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting? With perfectly fair scheduling, no waiting is necessary. Since no real-world scheduler is perfectly fair, we want minimal waiting. Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.

  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}. Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound. Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler. Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
\end{itemize}

This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks. (3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.

>>>>

\subsection{Decoupled Fallback Statistics}
To better understand how devices interact with \emph{DF}, we collect the following statistics during its execution:
\begin{itemize}
  \item \textbf{Spins per Workgroup:} How many times does a workgroup spin while waiting? With perfectly fair scheduling, no waiting is necessary. Since no real-world scheduler is perfectly fair, we want minimal waiting. Despite varying occupancy rates, the average spin rate across tested devices remains low (typically one or fewer), indicating that most workgroups experience minimal delays.

  \item \textbf{Total Successful Insertions:} The total number of fallback operations that successfully update a blocking (or late) tile to \emph{Ready}. Due to false-positive fallbacks, this count does not precisely represent the number of true starvations, but instead provides an upper bound. Ideally, this number would also be zero on devices with FPG, but in practice, it is again dependent on the fairness of the scheduler. Unlike the number of fallback initiations, this statistic is extremely low---typically well under 1\% even on devices without FPG---suggesting that true starvation is rare but catastrophic.
\end{itemize}
This data reveals three notable behaviors: (1) As demonstrated by the Apple M3, a scheduler can be fair yet still lack FPG, demonstrating that fairness and FPG are not necessarily correlated. (2) The ARM Mali-G78 MP20 exhibits the most unfair scheduling among all tested devices, with a fallback rate of 57.8\% relative to workgroups launched. However, the average number of spins per workgroup is only 2.52---below our fallback trigger threshold of 4 spins. This suggests a skewed distribution of waiting times, where workgroups either proceed with minimal delay or experience prolonged stalls that trigger fallbacks. (3) Despite this organically unfair scheduling, \emph{Decoupled Fallback} still achieves a 34.6\% speedup over \emph{RTS}, confirming its robustness to scheduler behavior in real-world conditions.

\subsection{Simulated Blocking}%
\label{sec:Sim-Blocking}
To further demonstrate the resilience of our technique to unfair schedulers, we simulate fallback rates significantly higher than those observed organically by intentionally omitting work tile reduction updates. Because these updates are skipped, the immediate successor workgroup reads \emph{Not Ready}, spinning and triggering a fallback. Starting with infrequent omissions---1 in 512 work tiles---we progressively increase the omission rate to near-continuous levels---1 in 2 work tiles---recording \emph{DF} statistics at each step.

\begin{figure*}
  \centering
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/fallbacksInitiated_plot.pdf}
    \caption{Total number of fallback operations triggered as simulated blocking rates increase. This statistic verifies that the simulation effectively induces fallbacks as intended.\label{fig:fallbacks_initiated}}
  \end{subfigure}\hfill
  \begin{subfigure}[t]{0.48\linewidth}
    \centering
    \includegraphics[width=\linewidth]{graphics/totalSpins_plot.pdf}
    \caption{Average number of waiting spins per workgroup as simulated blocking rates increase. Higher spin counts correlate with increased fallback operations, reinforcing the behavior observed in Figure~\ref{fig:fallbacks_initiated}.\label{fig:total_spins}}
  \end{subfigure}

@@
